Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-02-27 paper/paper.tex
Add the missing entry for nuprl
Add new overview

@@

    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & TODO \\
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}

>>>>

\yl{new overview starts here}
Here is an overview of our contribution.
\begin{itemize}
\item In Section~\ref{sec:spec},
we present the type system of \lang{},
a dependently typed system with an untyped conversion rule in the
style of Pure Type Systems~\citep{barendregt1991introduction}. The
\lang{} language supports many common features of a dependent
type theory: a countable universe hierarchy, a boolean base
type, an intensional identity type, and support for large elimination.
\item In Section~\ref{sec:logreldep}, we formulate logical consistency for \lang{}, the property of
interest, to motivate a logical relation, and proceed to define it as
an inductive relation which we later prove to be a partial
function. Based on the definition of the logical relation, we define
a notion of semantic typing and prove the fundamental theorem, from
which logical consistency follows as a corollary
(Section~\ref{sec:logrelproof}). Thanks to the good coverage of
features in our object language, our proof showcases the
special treatment required to model those features, thus making
our proof applicable to a broad range of type systems.
\item We strengthen our logical relation to prove the existence of
  $\beta$-normal forms (Section~\ref{sec:extension}) and
  $\beta\eta$-normal forms (Section~\ref{sec:betaeta}) for well-typed
  open \emph{and} closed terms. The modifications made to the logical
  relation are small and closely mirror an analogous extension for a
  simply typed language. We use these extensions to demonstrate the idea
  that once we have established the base techniques, we can port ideas
  that have been studied in the context of a simpler system to the
  dependently typed setting.
\item We mechanize all our proofs using the Coq proof assistant, with
  963 lines of code for the consistency proof and a moderate increase
  to 1357 lines of code for the normalization proof.
  We discuss the specifics
  related to our choice of Coq as our metatheory, including our use of
  off-the-shelf semantic engineering infrastructure and automation
  tools, in Section~\ref{sec:logrelmech}.
  Our proof scripts, with comments, are available to reviewers as
  supplementary materials. We use Coq's custom notations to ease the
  effort of matching the pen and paper representation to our formalism.
 \yl{I want to say something
  about the benefits of using existing libraries such as Autosubst 2
  and CoqHammer rather than rolling your own undocumented libraries that nobody
  want to (or can) reuse.}
Our development relies on off-the-shelf tools...
\item We discuss in Section~\ref{sec:relatedwork} how our proof
relates to existing proofs by logical relations and other proof
techniques for proving consistency and normalization. In
Section~\ref{sec:discuss}, we give an in-depth explanation on how
various design decisions (e.g. the choice of untyped equality) affect
the size of our proof and its extensibility to features such as
type-directed conversion.
\end{itemize}
We hope our success at creating a short and mechanized proof for a
relatively feature-complete language will encourage future researchers
to leverage the tool of logical relations in mechanized
proofs for dependent types.

\yl{new overview ends here}

@@

    % \hline
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat & T & {\boxedsymbols ✓} & Decidability of conversion \\
    % \hline
    MLTT-a-la-Coq & 1 & Id, Nat & T & {\boxedsymbols ✓} & Decidibility of type checking \\
    % \hline
    Core Nuprl & $\mathbb{N}$ & W-Types & E & {\boxedsymbols ✓} & Consistency \\
    \hline
    % CIC & $\mathbb{I}$,$\mathbb{N}$ & General & U & {\boxedsymbols ✓}\\
    % CoreSpec & \\
    % PCUIC & \\
    %   & \lang{} & \citet{anand2014towards} &
    %                                          \citet{barras2012semantical}
    % & \citet{martin-lof-a-la-coq}  \\
    % \hline
    % Universe & Countable & Countable & Impredicative + Countable & One\\
    % % \hline
    % % Large Elimination & {\boxedsymbols ✓} & {\boxedsymbols ✓} & {\boxedsymbols ✓} \\
    % \hline
    % Inductive Types & Id, Bool & W types & General inductive types & Nat \\
    % \hline
    % Convertibility & Untyped & Untyped,Extensional & Untyped & Typed \\
  \end{tabular}
